---
title: 'class12: Homework'
author: 'Leah Johnson, PID: A17394690'
date: "2026-02-15"
output:
  pdf_document: default
  html_document: default
toc: true
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file:

```{r}
mxl <- read.csv("SampleGenotypes.csv")
head(mxl)
```


```{r}
table(mxl$Genotype..forward.strand.)
```


```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("SampleGenotypesGBR.csv")
head(gbr)
```


Find proportion of G|G

```{r}
table(gbr$Genotype..forward.strand.)
```


```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.




# Section 4: Population Scale Analysis [HOMEWORK]


> **Q13**: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.


How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```


There are 462 total individual samples in this data-set.

```{r}
nrow(expr)
```


There are 108 samples for A/A genotype, 233 samples for A/G genotype, and 121 samples for G/G genotype.

```{r}
table(expr$geno)
```


What are the median expression levels for each genotype?


A/A median is 31.25.

```{r}
summary(expr$exp[expr$geno == "A/A"])
```


A/G median is 25.07.

```{r}
summary(expr$exp[expr$geno == "A/G"])
```


G/G median is 20.07.

```{r}
summary(expr$exp[expr$geno == "G/G"])
```




> **Q14**: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?



By looking at the plot, I can infer that there is less genome expression with the G/G genotype, as the calculated median expression is ~20.07 which is less than the calculated median genome expression with A/A genotype at ~31.82. Therefore, the SNP does effect the expression of ORMDL3 because the median genome expression levels differ by genotype differences amongst all samples considered in this data-set.



```{r}
library(ggplot2)

```


Let's make a boxplot:

```{r}
genome_ggplot <- ggplot(expr) + aes(x=geno, y=exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```


```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) + 
  geom_boxplot(notch=TRUE)
```









