---
title: "class05:DataVisualization_GGPLOT"
author: "Leah Johnson"
format: pdf
toc: true
---

## Background

There are lots of ways to make figures in R. These include so-called "base R" graphics (e.g. 'plot()') and tons of add-on packages like **ggplot2**.

For example, here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command install.packages().

> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want!

Every time we want to use one of these "add-on" packages we need to load it up in R with the 'library()' function:

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```


Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thetics, how the data maps to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point()
```


```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  labs(title="Speed and Stopping Distances of Cars",
       x="Speed (mph)",
       y="Stopping Distance (ft)",
       subtitle="Data from the 'cars' object",
       caption="Dataset: 'cars'") + 
  geom_smooth(method="lm", se=FALSE, col="cyan4") + 
  theme_bw()
```

render it out
```{r}
p
```

```{r}
p + theme_minimal()
```


## Gene expression plot

We can read the input data from the class website.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
```
```{r}
colnames(genes)
ncol(genes)
```
```{r}
table(genes$State)
```
```{r}
round(table(genes$State)/nrow(genes) * 100, 2)
```



Version 1 plot:

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) + 
  geom_point()
```

Version 2 plot: Let's color by State so we can see the "up" and "down" significant genes compared to the "unchanging" genes.

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point()
```

Version 3 plot: Let's modify the default colors to something we like.

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() + 
  scale_colour_manual(values=c("darkgoldenrod2", "orchid", "red")) + 
  labs(x="Control (no drug)",
       y="Drug",
       title="Gene Expression Changes Upon GLP-1") + 
  theme_bw()
```

## Going Further

Let's have a look at the famous **gapminder**  
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

```{r}
ggplot(data=gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point(alpha=0.3)
```

Let's facet (i.e. make a separate plot) by continent rather than the hot mess above.

```{r}
ggplot(data=gapminder) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point(alpha=0.3) + 
  facet_wrap(~continent)
```


## Custom Plots

How large is this gapminder dataset()

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up...
'install.packages(dplyr)' and then 'library(dplyr)'

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```


```{r}
filter(gapminder_2007, country=="Ireland")
```


```{r}
filter(gapminder, year==2007, country=="Ireland")
```


```{r}
filter(gapminder_2007, country=="United States")
```


> Q. Make a plot comparing 1977 and 2007 for all countries.

```{r}
input <- filter(gapminder, year %in% c(1977, 2007))
ggplot(input) + 
  aes(x=gdpPercap, y=lifeExp, col=continent) + 
  geom_point(alpha=0.5) + 
  facet_wrap(~year)
```



