---
title: "class18_PertussisProject"
author: "Leah Johnson, PID: A17394690"
format: pdf
toc: true
---

# Background

Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone, but is most deadly for infants (under 1 years old).

# CDC tracking data

The CDC tracks the number of Pertussis cases.

```{r, echo=FALSE}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
)
```


> **Q1.** With the help of the R “Addin” package 'datapasta()' assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line()
```

> **Q2.** Add annotation lines for the milestones of wP vaccination roll-out (1946) and the switch to aP vaccine (1996).

After introduction of the wP vaccine in 1946, the number of Pertussis cases dropped, but after the introduction of the aP vaccine in 1996, the number of Pertussis cases began to rise again.

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 1946, col="blue", lty=2) +
  geom_vline(xintercept = 1996, col="red", lty=2) + 
  geom_vline(xintercept = 2020, col="darkgreen", lty=2)
```

> **Q3.** Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine in 1996, the number of Pertussis cases increases after the brief lag visible in the graph above. These results could be explained by numerous factors, such as new technology being used in hospitals to detect Pertussis virus, less people getting the vaccine to begin with, and/or 'waning efficacy' of the aP vaccine.


# Exploring CMI-PB Data

The CMI-PB project's < https://www.cmi-pb.org/ > 
"The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination."

Make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection.

They make their data available on JSON format API. We can read this into R with the 'read.json()' function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject",
                     simplifyVector = TRUE)
head(subject)
```


> **Q4.** How many aP and wP individuals are there in this 'subject' table?

```{r}
subject$infancy_vac
```

```{r}
table(subject$infancy_vac)
```


> **Q5.** How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


> **Q6.** What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```


> **Q6a.** Is this representative of the US population?

The White and Asian Male and Female groups have the greatest counts of those who received the vaccination.


## Working with Dates

```{r}
library(lubridate)

## What is today's date?
today()

## How many days have passed since new year 2000?
today() - ymd("2000-01-01")

## What is this in years?
time_length( today() - ymd("2000-01-01"),  "years")
```


> **Q7.** Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

(i) Average age of wP individuals:

```{r}
library(dplyr)

wp <- subject %>% filter(infancy_vac == "wP")
```

```{r}
wp$age <- today() - ymd(wp$year_of_birth)

mean(wp$age)
```

```{r}
library(dplyr)

wp_age <- round( summary( time_length( wp$age, "years" ) ) )
```

(ii) Average age of aP individuals:

```{r}
ap <- subject %>% filter(infancy_vac == "aP")
```

```{r}
ap$age <- today() - ymd(ap$year_of_birth)

mean(ap$age)
```

```{r}
library(dplyr)

ap_age <- round( summary( time_length( ap$age, "years" ) ) )
```


> **Q8.** Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)

age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


> **Q9.** With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(age_at_boost, fill=infancy_vac) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years") + 
  ylab("Count")
```

Based on the spread of data seen in the histograms above, the aP and wP groups differ in the ages of all individuals at the time of the boost. The mean of individuals for aP is approximately 20 years old and the mean age of individuals for wP is approximately 32 years old. The p-value suggests the differenced between ages of indivuals at the time of the boost for aP and wP is significant: 

```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```


We can read more tables from the CMI-PB database:

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To make sense of this data, we need to "join" (a.k.a "merge" or "link") all these tables together. Only then will you know that a given anti-body measurement (from the 'ab_titer' table) was collected on a certain date (from the 'specimen' table) from a certain wP or aP subject (from the 'subject' table).

We can use **dplyr** and the '*_join()' family of functions to do this.


> **Q9.** Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```


> **Q10.** Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Let's do one more 'inner_join()' to join the 'ab_titer' with all our 'meta' data.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```


> **Q11.** How many different Ab "isotype" values are in this dataset?

```{r}
table(abdata$isotype)
```


> **Q11a.** How many different "antigen" values are measured?

```{r}
table(abdata$antigen)
```


> **Q12.** How many different "dataset" values are measured?

```{r}
table(abdata$dataset)
```


# Examine IgG ab_titer levels


Let's focus on IgG isotype:
("|>" or "%>%")

```{r}
igg <- abdata |>
  filter(isotype=="IgG")

head(igg)
```


> **Q13.** Make a summary boxplot of ab_titer levels (MFI) for all antigens:

Make a plot of 'MFI_normalised' values for all 'antigen' values.

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```


> **Q14.** What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens "PT", "FIM2/3", and "FHA" appear to have the widest range of ab titer values. The whole-cell (wP) vaccine has [more of] all these elements.


> **Q14a.** Is there a difference for these responses between aP and wP individuals?

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col=infancy_vac) + 
  geom_boxplot() + 
  facet_wrap(~infancy_vac)
```


> **Q14b.** Is there a difference with time (i.e. before booster shot vs. after booster shot)?

The antibody levels peak in graphs 6 and 7 and then decrease.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

Another version:

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```


> **Q15.** Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


> **Q16.** What do you notice about these two antigens time courses and the PT data in particular?

The PT, FIM2/3, and OVA increase over time, with PT exceeding the OVA values. The PT values peak in the 5-8 region and then decrease. 


> **Q17.** Do you see any clear difference in aP vs. wP responses?

The trend described in Q16 pertains to both the aP and wP response values.


> **Q18.** Does this trend look similar for the 2020 dataset?

Generally yes, despite a couple aP and wP outliers present in the 2020 dataset.

```{r, warning=FALSE}
## filter to 2021 dataset, IgG and PT only
ab.PT.21 <- abdata %>%
  filter(dataset == "2021_dataset",
         isotype == "IgG",
         antigen == "PT")

ggplot(ab.PT.21) +
  aes(x=planned_day_relative_to_boost,
      y=MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point(size=1) +
    geom_line(alpha=0.3) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  geom_smooth(aes(x=planned_day_relative_to_boost, y=MFI_normalised, group = infancy_vac), method = "loess", span=0.2, se = FALSE, linewidth = 1.5) +
  labs(title="CMI-PB 2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") + 
    xlab("Day Relative to Boost") + 
    ylab("Normalised MFI")
```

```{r, warning=FALSE}
## filter to 2020 dataset, IgG and PT only
ab.PT.20 <- abdata %>%
  filter(dataset == "2020_dataset",
         isotype == "IgG",
         antigen == "PT")

ggplot(ab.PT.20) +
  aes(x=planned_day_relative_to_boost,
      y=MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point(size=1) +
    geom_line(alpha=0.3) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  geom_smooth(aes(x=planned_day_relative_to_boost, y=MFI_normalised, group = infancy_vac), method = "loess", span=0.2, se = FALSE, linewidth = 1.5) +
  labs(title="CMI-PB 2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") + 
    xlab("Day Relative to Boost") + 
    ylab("Normalised MFI")
```


# Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```


Join 'rna' with 'meta'
```{r}
ssrna <- inner_join(rna, meta)
```


> **Q19.** Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```


> **Q20.** What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The maximum expression of this gene is around visit 4.


> **Q21.** Does this pattern in time match the trend of antibody titer data? If not, why not?

The max expression of this IGHG1 gene occurs earlier than the max expression of the previously explored antigens. However, the timing of max gene expression between the IGHG1 gene and other mentioned antigens is very close. This could be explained by the longer amount of time it takes for cells to produce antigens, and for antigen expression to, therefore, occur, in contrast to gene expression.


```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```


No difference in wP and aP, even if we focus on a particular visit.

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```


