---
title: "class07: Machine Learning 1"
author: "Leah Johnson, PID: A17394690"
format: pdf
toc: true
---

## Background

Today we will begin exploration of important machine learning methods with a focus on **clustering** and **dimensionality reduction**.

To start testing these methods, let's make up some sample data to cluster where we know what the answer should be.


```{r}
hist( rnorm(3000, mean=10) )
```

> Q. Can you generate 30 numbers centered at +3 taken at random from a normal distribution? 

```{r}
tmp <- c(rnorm(30, mean=3), rnorm(30, mean=-3))

x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

## K-means clustering

The main function in "base R" for K-means clustering is called 'kmeans()', let's try it out:

```{r}
k <- kmeans(x, centers=2)
k
```

> Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

> Q. What component of your kmeans result object has the cluster size (i.e. how many points in each cluster?)

```{r}
k$size
```

> Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster)?

```{r}
k$cluster
```

> Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15, cex=2)
```

> Q. Can you run 'kmeans()' again, cluster 'x' into 4 clusters and plot the results just like you did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k <- kmeans(x, centers=4)
k
```

```{r}
k$centers
```

```{r}
k$size
```

```{r}
k$cluster
```

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=15, cex=2)
```

> **Key-point** K-means will always return the clustering that we ask for (this is the "k" or "centers" in k-means)

```{r}
k$tot.withinss
```

## Hierarchical clustering

The main function for hierarchical clustering in base R is called 'hclust()'. One of the main differences with respect to the 'kmeans()' function is that you can not just pass your input data directly to 'hclust()' - it needs a "distance matrix" as input. We can get this from lot's of places including the 'dist()' function.

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

We can "cut" the dendrogram or "tree" at a given height to yield our "clusters". For this, we use the function 'cutree()'.

```{r}
plot(hc)
abline(h=10, col="red")
grps <- cutree(hc, h=10)
```

```{r}
grps
```


> Q. Plot our data 'x' colored by the clustering result from 'hclust()'?

```{r}
grps <- cutree(hc, h=10)
plot(x, col=grps)
```

```{r}
plot(hc)
abline(h=5, col="red")
grps2 <- cutree(hc, h=5)
```

```{r}
grps2
```

```{r}
grps2 <- cutree(hc, h=5)
plot(x, col=grps2)
```

## Principal Component Analysis (PCA)

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics.

## PCA of UK food data

Read data on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

It looks like the row names are not set properly. We can fix this:

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
x
```

A better way to do this is fix the row names assignment at import time:

```{r}
x <- read.csv(url, row.names = 1)
```


> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second approach to solving the 'row-names problem' mentioned above because it is more clear and concise.

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside=FALSE, col=rainbow(nrow(x)))
```

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The code compares 4 different countries (2 at a time), England, Wales, Scotland, and N.Ireland, and each point represents one of the 17 'food' groups we are concerned about. If the given point lies on the diagonal for a given plot, it means there are the same or similar  consumption values of a specific food group for both compared countries.

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

## Heatmap

We can install the **pheatmap** package with the 'install.packages()' command that we used previously. Remember that we always run this in the console and not a code chunk in our quarto document/

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

Of all these plot really only the 'pairs()' plot was useful. This however took a bit of work to interpret and will not scale when I am looking at larger datasets.

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Based on the pairs and heatmap figures, Whales and England cluster most closely together, suggesting their food consumption patterns are more similar. The dendrogram on the top of the heatmap presents a closer clustering between Whales and England in comparison to Whales and N. Ireland, for example. Knowing that the top dendrogram represents how countries cluster together and the left dendrogram represents how foods cluster together, along with the shading of the heatmap, I can eadily tell what the main differences are between N. Ireland and the other countries.

## PCA to the rescue

```{r}
library("ggplot2")
```

The main function in "base R" for PCA is called 'prcomp()'.

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

> Q. How much variance is captured in the first PC?

67.4% variance is captured in PC1.


> Q. How many PCs do I need to capture 90% of the total variance in the dataset?

Two PCs capture 96.5% of the total variance.


> Q7. Plot our main PCA result. Folks can call this different things like "PC plot", "ordienation plot", "score plot", "PC1 vs PC2 plot", etc.

```{r}
df <- as.data.frame(pca$x)
df$Country <- rownames(df)
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 2) + 
  geom_text(vjust = -0.5)
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```


```{r}
attributes(pca)
```

To generate our PCA score plot we want the 'pca$x' component of the result object.

```{r}
pca$x
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)
df$Country <- rownames(df)
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) + 
  geom_point(col=my_cols, size = 2) + 
   geom_text(vjust = -0.5)
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

## Digging deeper (variable loadings)

How do the original variables (i.e. the 17 different foods) contribute to our new PCs?

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```

This loading plot reduced the dimensions even further of the data from the PC1 loading plot and the PC2 loading plot displays the variation in the "largest positive loading scores" with the uppermost "soft drinks."















